Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2015-09-18 Univalence/response.txt

We thank the reviewers for their comments.

A general theme of the reviews appears to request a better presentation of the theorems and how they all relate to each other. A compressed and intuitive presentation would be as follows. First we have three independent constructions:

  A : Set     <------- HoTT-style equivalences -------> B : Set

  A : Fin m <----------- Permutations ------------> B : Fin n

  A : Pi       <----------- Pi-combinators ----------> B : Pi

Each construction relates a family of types via appropriate relations; each construction has a rig structure (additive and multiplicative units, commutative and associative sums and products, and distributivity of products over sums). We then have a second level:

  e : HoTT-Equivalence <----------- Eq. of equivalences --------> f : HoTT-Equivalence

  p : Permutations         <----------- Eq. of permutations ------->  q : Permutations

  c : Pi-combinators     <------------- Pi-optimizations --------> d : Pi-combinators

We then have theorems that connect some of the constructions vertically allowing us to move across representations.

This are the "raw" results without interpretation, context, or applications. Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT and univalence and with practical connections to reversible circuits and their optimizations. The connections are there and we are important to appreciate the significance of the theorems but the presentation could indeed be better.

### 2015-09-18 Univalence/response.txt

  A : Set     <------- HoTT-style equivalences -------> B : Set

  A : Fin m <----------- Permutations ------------> B : Fin n

  A : Pi       <----------- Pi-combinators ----------> B : Pi

Each construction relates a particular representation of types via appropriate relations; each construction has a rig structure (additive and multiplicative units, commutative and associative sums and products, and distributivity of products over sums). We then have a second level:

@@

  e : HoTT-Equivalence <----------- Eq. of equivalences --------> f : HoTT-Equivalence

  p : Permutations         <----------- Eq. of permutations ------->  q : Permutations

  c : Pi-combinators     <------------- Pi-optimizations --------> d : Pi-combinators

The two levels of constructions are connected via appropriate categories. For example, we have a category whose objects are Pi types, whose morphisms are Pi-combinators, and whose equality of morphisms if Pi-optimizations and this category is a (free) symmetric rig groupoid. We finally have theorems that connect some of the constructions vertically allowing us to move across representations.

This are the "raw" results without interpretation, context, or applications. Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT, with Agda formalization concerns, and with practical connections to reversible circuits and their optimizations. We acknowledge that these connections are somewhat distracting and that the presentation could be much improved but these connections are all critical and essential. Indeed some of the connections above are trivial in "conventional mathematics" but some of that triviality hides implicit isomorphisms (changes of representations) that are important computationally.  Our construction shows, for example, that transporting properties across equivalences of finite types can be done via executing permutations, something which has a clear computational cost and which itself depends on the choice of representations of the permutations. Since this is POPL, the presentation is also biased towards turning the constructions into actual executable programming languages with executable optimizers and with an eye towards higher-order extensions. To our knowledge, there are no programming languages for representing --and reasoning about-- a rich notion of "permutations" (including sums and products). Furthermore, the higher-order extensions are crucial for abstractions over such permutations by allowing the permutations themselves to be treated as data that can be constructed and examined dynamically.

### 2015-09-19 Univalence/response.txt

  c : Pi-combinators     <------------- Pi-optimizations --------> d : Pi-combinators

The two levels of constructions are connected via appropriate categories. For example, we have a category whose objects are Pi types, whose morphisms are Pi-combinators, and whose equality of morphisms if Pi-optimizations and this category is a (free) symmetric rig groupoid. We finally have theorems that connect some of the constructions vertically allowing us to move across representations.

This are the "raw" results without interpretation, context, or applications. Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT, with Agda formalization concerns, and with practical connections to reversible circuits and their optimizations. We acknowledge that these connections are somewhat distracting and that the presentation could be much improved but these connections are all critical and essential. Indeed some of the connections above are trivial in "conventional mathematics" but some of that triviality hides implicit isomorphisms (changes of representations) that are important computationally.  Our construction shows, for example, that transporting properties across equivalences of finite types can be done via executing permutations, something which has a clear computational cost and which itself depends on the choice of representations of the permutations. Since this is POPL, the presentation is also biased towards turning the constructions into actual executable programming languages with executable optimizers and with an eye towards higher-order extensions. To our knowledge, there are no programming languages for representing --and reasoning about-- a rich notion of "permutations" (including sums and products). Furthermore, the higher-order extensions are crucial for abstractions over such permutations by allowing the permutations themselves to be treated as data that can be constructed and examined dynamically.

Specific (and short with apologies) responses:

* I think our excitement with all the connections apparently left a "hint of arrogance." We sincerely apologize for that.

* The supplementary material gives one small example of rewriting circuits using the 2-axioms. We do not have larger experiments at this point but we have done enough to notice that the rewriting of one circuit to another does not necessarily take an exponential number of steps: the diagram on p.10 shows an equivalence that is independent of the sizes of c1 and c2 and the sizes of A, B, C, and D.

* Our axioms vs. Lapalaza diagrams: In some sense they are indeed "the same". The general process is to give judicious names to each side of each diagram and then proceed to prove equivalence of the two expressions. In some cases this was achieved by adding a 2-axiom that directly relates the terms but in others it was possible to simplify the two sides using other 2-axioms. The process led us to refine the original 1-axioms of Pi which implicitly collapsed certain terms that should only been related by explicit non-trivial paths.

* Proofs: All the theorems (except Thm 2) are in our Agda formalization. This code was not presentable as an artifact at the time of submission and required some serious refactoring which is under progress. Thm 2 is indeed wrong was the informal statement was included by mistake (but it is subsumed by Thm 9).

* The 3rd axiom on p9 says c;(!c)=id which proves that swap;swap=id. (The bottom of p4 notes that every combinator has a dual according to figure 1 with swap being its own inverse.) The existence of inverses for 1-combinators is a meta-property (not a constructor of the inductive datatype of combinators but a function mapping combinators to combinators). The same applies to inverses for 2-combinators.

* Completeness in Thm 11. *** BRIEFLY EXPLAIN HOW WE GET COMPLETENESS FROM LAPLAZA ***

* *** COMMENT ON PHONY MULTIPLICATION ***

### 2015-09-19 Univalence/response.txt

A general theme of the reviews appears to request a better presentation of the
theorems and how they all relate to each other. A compressed and intuitive
presentation would be as follows. First we have three independent
constructions:

  A : Set    <------- HoTT-style equivalences -------> B : Set

  A : Fin m  <------------ Permutations -------------> B : Fin n

  A : Pi     <------------ Pi-combinators -----------> B : Pi

@@

Each construction relates a particular representation of types via appropriate
relations; each construction has a rig structure (additive and multiplicative
units, commutative and associative sums and products, and distributivity of
products over sums). We then have a second level:

  e : HoTT-Equivalence <--------- Equiv. of equivalences --------> f : HoTT-Equivalence

  p : Permutations     <--------- Equiv. of permutations --------> q : Permutations

  c : Pi-combinators   <-------------- Pi-optimizations ---------> d : Pi-combinators

The two levels of constructions are connected, and form categories. For
example, we have a category whose objects are Pi types, whose morphisms are
Pi-combinators, and whose equality of morphisms is Pi-optimizations; this
category is a (free) symmetric rig groupoid. We finally have theorems that
connect some of the constructions vertically allowing us to move across
representations.

These are the "raw" results without interpretation, context, or applications.
Perhaps unfortunately, the current presentation of the paper mixes these "raw"
results with theoretical connections to HoTT, with Agda formalization concerns,
and with practical connections to reversible circuits and their optimizations.
We acknowledge that these connections can be perceived as distracting, and that
their presentation could be much improved (which we are working on), BUT these
connections are all critical and essential -- and are why this is being sent to
POPL rather than (say) Theory and Applications of Category Theory.

Indeed some of the connections above are trivial in "conventional mathematics";
however some of those same connections hide implicit isomorphisms (changes of
representations) that are important computationally.  Our construction shows,
for example, that transporting properties across equivalences of finite types
can be done via executing permutations, something which has a clear
computational cost and which itself depends on the choice of representations of
the permutations. Since this is POPL, the presentation emphasizes turning these
constructions into actual executable programming constructs of a programming
language, complete with executable optimizers, and with an eye towards
higher-order extensions. To our knowledge, there are no programming languages
for representing --and reasoning about-- a rich notion of "permutations"
(including sums and products).  Furthermore, the higher-order extensions are
crucial for abstractions over such permutations by allowing the permutations
themselves to be treated as data that can be constructed and examined
dynamically.

@@

Specific (and short with apologies) responses:

* I think our excitement with all the connections apparently left a "hint of
  arrogance." We sincerely apologize for that, and will fix.

@@

* The supplementary material gives one small example of rewriting circuits
  using the 2-axioms. We do not have larger experiments at this point but we
  have done enough to notice that the rewriting of one circuit to another does
  not necessarily take an exponential number of steps: the diagram on p.10
  shows an equivalence that is independent of the sizes of c1 and c2 and the
  sizes of A, B, C, and D.

* Our axioms vs. Laplaza diagrams: Yes, the level 2 combinators for 
  distributivity properties and the interaction of combinators involving 0
  are the same.  We apparently were not clear enough about what changed:

  The general process of incorporating these diagrams as combinators is to give
  judicious names to each natural transformation of each diagram representing
  a natural isomorphism, and then proceed to prove equivalence of the two
  expressions (for all 3 constructions). In some cases this was achieved by
  adding a 2-axiom that directly relates the terms but in others it was possible
  to simplify the two sides using other 2-axioms. The process led us to refine
  the original 1-axioms of Pi which implicitly collapsed certain terms that
  should only been related by explicit non-trivial paths. Thus the most 
  apparent changes were to the programming language Pi itself, and not the
  combinators derived from Laplaza's diagrams.  In other words, we let the
  ``proper'' categorical semantics guide us to define the syntax of a new
  version of Pi which had better meta-properties.

* Proofs: All the theorems (except Thm 2) are in our Agda formalization. This
  code was not presentable as an artifact at the time of submission and
  required some serious refactoring which is under progress. Thm 2 is indeed
  wrong and the informal statement was included by mistake; the core of what
  we meant to convey is in the (correct) Thm 9.

### 2015-09-19 Univalence/response.txt

  c : Pi-combinators   <-------------- Pi-optimizations ---------> d : Pi-combinators

The two levels of constructions are connected, and form categories. For
example, we have a category whose objects are Pi types, whose morphisms are
Pi-combinators, and whose equality of morphisms is Pi-optimizations; this
category is a (free) symmetric rig groupoid. We finally have theorems that
connect some of the constructions vertically allowing us to move across
representations.

These are the "raw" results without interpretation, context, or applications.
Perhaps unfortunately, the current presentation of the paper mixes these "raw"
results with theoretical connections to HoTT, with Agda formalization concerns,
and with practical connections to reversible circuits and their optimizations.
We acknowledge that these connections can be perceived as distracting, and that
their presentation could be much improved (which we are working on), BUT these
connections are all critical and essential -- and are why this is being sent to
POPL rather than (say) Theory and Applications of Category Theory.

Indeed some of the connections above are trivial in "conventional mathematics";
however some of those same connections hide implicit isomorphisms (changes of
representations) that are important computationally.  Our construction shows,
for example, that transporting properties across equivalences of finite types
can be done via executing permutations, something which has a clear
computational cost and which itself depends on the choice of representations of
the permutations. Since this is POPL, the presentation emphasizes turning these
constructions into actual executable programming constructs of a programming
language, complete with executable optimizer, and with an eye towards
higher-order extensions. To our knowledge, there are no programming languages
for representing --and reasoning about-- a rich notion of "permutations"
(including sums and products).  Furthermore, the higher-order extensions are
crucial for abstractions over such permutations by allowing the permutations
themselves to be treated as data that can be constructed and examined
dynamically.

@@

* The supplementary material gives one small example of rewriting circuits
  using the 2-axioms. We do not have larger experiments at this point but we
  have done enough to notice that the rewriting of one circuit to another does
  not necessarily take an exponential number of steps: the diagram on p.10
  shows an equivalence that is independent of the sizes of c1 and c2 and the
  sizes of A, B, C, and D.

* Our axioms vs. Laplaza diagrams: Yes, the level 2 combinators for 
  distributivity properties and the interaction of combinators involving 0
  are the same.  We apparently were not clear enough about what changed:

  The general process of incorporating these diagrams as combinators is to give
  judicious names to each natural transformation of each diagram representing
  a natural isomorphism, and then proceed to prove equivalence of the two
  expressions (for all 3 constructions). In some cases this was achieved by
  adding a 2-axiom that directly relates the terms but in others it was possible
  to simplify the two sides using other 2-axioms. The process led us to refine
  the original 1-axioms of Pi which implicitly collapsed certain terms that
  should only been related by explicit non-trivial paths. Thus the most 
  apparent changes were to the programming language Pi itself, and not the
  combinators derived from Laplaza's diagrams.  In other words, we let the
  ``proper'' categorical semantics guide us to define the syntax of a new
  version of Pi which had better meta-properties.

  So it is not the coherences which are more "small step", it is the 
  underlying programming language.

@@

* Proofs: All the theorems (except Thm 2) are in our Agda formalization. This
  code was not presentable as an artifact at the time of submission and
  required some serious refactoring which is under progress. Thm 2 is indeed
  wrong and the informal statement was included by mistake; the core of what
  we meant to convey is in the (correct) Thm 9.

* The 3rd axiom on p9 says c;(!c)=id which proves that swap;swap=id. (The
  bottom of p4 notes that every combinator has a dual according to figure 1
  with swap being its own inverse.) The existence of inverses for 1-combinators
  is a meta-property (not a constructor of the inductive datatype of
  combinators but a function mapping combinators to combinators). The same
  applies to inverses for 2-combinators.

* Completeness in Thm 11. *** BRIEFLY EXPLAIN HOW WE GET COMPLETENESS FROM LAPLAZA ***

* lifting of the 0d combinators: The problem is when given a : T1 <=> T2,
  b : T3 <=> T4, to construct something of type T1 x T3 <=> T2 x T4.  If
  b is the identity, yes it works.  But it does not work in general.
  There is even an (easy!) general definition for what it ought to be.
  The point is that that definition is not functorial.  If it were, that
  would contradiction Thomasson's result.

* We would like to remark that Thomasson believed strongly enough that a 
  particular construction (which had been published several times) was in 
  fact wrong, that he published a paper point exactly that, and only that.
  Most PL theorists would not understand that this result actually applies to
  constructions they are familiar with: the Int-construction, which applies so
  nicely to language built atop linear logic.  We believe, just as strongly 
  as Thomasson, that pointing out that the ``obvious'' generalization to
  the setting with two monoidal constructions (such as ours) does not work,
  for the same reasons.  Yes, we do realize that including a negative result
  in paper makes it harder to publish; we nevertheless feel that we can save
  others a lot of fruitless pain by attempting to do so.

* Referee 4 thinks that "The main result of the paper is that the category of
  finite sets and bijections, with the two monoidal structures given by sum and
  product, is the free rig groupoid." And then further that " A serious
  weakness is that the main result has no proof, nor even a proof sketch."  This
  would be because we did not actually claim this result as a theorem!  [We
  believe it to be true, but as of this writing, that proof is still 
  incomplete.]
  Our main result is that weak rig categories are exactly the setting for the
  semantics for a programming language of reversible circuits.  Furthermore, the
  coherence laws of such a category exactly give complete rules for deciding
  equivalence of such circuits.  In other words, theorems 10 and 11.  We also
  consider theorem 9 (that *all* types, not just finite ones, equivalences and
  equivalences-of-equivalences), form a weak rig groupoid to be quite important
  for PL theory.

### 2015-09-20 Univalence/response.txt

Specific responses:

* Proofs: All the theorems (except Thm 2 and Thm 11) are in our Agda
  formalization. This code was not presentable as an artifact at the
  time of submission and required some serious refactoring which is
  under progress. Thm 2 is indeed wrong and the informal statement was
  included by mistake; the core of what we meant to convey is in the
  (correct) Thm 9. For Thm 11, the first part was proved in previous
  work; the second part follows from the logical interpretation of
  coherence in the categorical sense as completeness of a syntactic
  axiomatization (Pi axioms) with respect to a model (rig category of
  type equivalences) (see e.g.,
  http://www.mi.sanu.ac.rs/~kosta/coh.pdf)

* lifting of the 0d combinators: The problem is when given a : T1 <=>
  T2, b : T3 <=> T4, to construct something of type T1 x T3 <=> T2 x
  T4.  If b is the identity, yes it works.  But it does not work in
  general.  There is even an (easy!) general definition for what it
  ought to be.  The point is that that definition is not functorial.
  If it were, that would contradiction Thomasson's result. Regarding
  the construction given at the end of the report, that part is right,
  it is the 'next' step which breaks down.  In more detail

  T1 = a - b
  T2 = a' - b'
  T3 = c - d
  T4 = c' - d'

  a : T1 <=> T2
     : a + b' <-> a' + b

  a' : T1 x T3 <=> T2 x T3
     : (a-b) x (c-d) <=> (a'-b') x (c-d)
     : (ac+bd - bc+ad) <=> (a'c+b'd - b'c+a'd)
     : ac+bd+b'c+a'd <-> a'c+b'd+bc+ad

  (same as in the given proof, just flipped)

  b : T3 <=> T4
  b : c + d' <-> c' + d

  b' : T2 x T3 <=> T2 x T4
      : (a'-b') x (c-d) <=> (a'-b') x (c'-d')
      : (a'c+b'd - b'c+a'd) <=> (a'c'+b'd' - b'c'+a'd')
      : a'c'+b'd+b'c'+a'd' <-> a'c'+b'd'+b'c+a'd

  Now a' and b' do not compose.

* Referee 4 states that "The main result of the paper is that the
  category of finite sets and bijections, with the two monoidal
  structures given by sum and product, is the free rig groupoid." And
  then further that " A serious weakness is that the main result has
  no proof, nor even a proof sketch."  This would be because we did
  not actually claim this result as a theorem!  [We believe it to be
  true, but as of this writing, that proof is still incomplete.]  Our
  main result is that weak rig categories are exactly the setting for
  the semantics for a programming language of reversible circuits.
  Furthermore, the coherence laws of such a category exactly give
  complete rules for deciding equivalence of such circuits.  In other
  words, theorems 10 and 11.  We also consider theorem 9 (that *all*
  types, not just finite ones, equivalences and
  equivalences-of-equivalences), form a weak rig groupoid to be quite
  important for PL theory.

-- Optional general comment

### 2017-10-16 survey/outline.tex
outline

\documentclass{article}
\usepackage{fullpage}
\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
\author{Jacques Carette, Roshan James, Amr Sabry}
\begin{document}
\maketitle

@@

@@

Computation in the ``real world'' vs. idealized computation in abstract models

@@

Conflict between Church-Turing thesis, textbook quantum mechanics, and RSA.

Conflict between quantum mechanics and logic.

@@

Pi languages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Application: Information Flow Security}

??

>>>>

\documentclass[12pt]{article}
\usepackage{fullpage}
\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
\author{Jacques Carette, Roshan James, Amr Sabry}
\begin{document}
\maketitle

@@

Abstract models of computation: Turing machine, RAM, Von Neumann
architecture, $\lambda$-calculus, logic, discrete mathematics

Cyber-physical systems increasingly common. Are computing foundations
adequate?

Toffoli 1980: Mathematical models of computation are abstract
constructions, by their nature unfettered by physical laws. However,
if these models are to give indications that are relevant to concrete
computing, they must somehow capture, albeit in a selective and
stylized way, certain general physical restrictions to which all
concrete computing processes are subjected.

@@

Computation in the ``real world'' vs. idealized computation in
abstract models. 
\begin{itemize}
\item Mathematical abstraction of sorting is well-understood; sorting
  petabytes of data violates many of the assumptions: hardware
  component failure frequent, global synchronization impossible;
  consistency must be eventual; etc. 
\item Mathematical abstraction of sortedness property: easy; in the
  real world with petabytes of evolving data; it is not clear that it
  is even possible to check sortedness; easy to get inconsistent
  results
\item Noise does not exist in our mathematical abstractions of
  computation; essential use in differential privacy
\item Mathematical models of secure computing typically ignore power
  consumption, electromagnetic signatures; actual attacks based on DPA
  and DEMA however
\end{itemize}

@@

Conflict between Church-Turing thesis, textbook quantum mechanics, and
RSA: Either Shor’s algorithm is not ``natural''. (Textbook quantum
mechanics is wrong); or, Shor’s algorithm is ``natural'' and there is
no classical counterpart. (There are ``natural'' computing models that
are exponentially faster than the Turing Machine); or, there is an
efficient classical factoring algorithm. (Possible but surprising); At
least one of these wild claims is true!!!

Conflict between quantum mechanics and logic. Possibilities:

\paragraph*{Revise quantum mechanics}.

The mathematician's vision of an unlimited sequence of totally
reliable operations is unlikely to be implementable in this real
universe.  But the real world is unlikely to supply us with unlimited
memory or unlimited Turing machine tapes. Therefore, continuum
mathematics is not executable, and physical laws which invoke that can
not really be satisfactory. They are references to illusionary
procedures.  (Landauer 1996 and 1999)

I want to talk about the possibility that there is to be an exact
simulation, that the computer will do exactly the same as nature. If
this is to be proved and the type of computer is as I've already
explained, then it's going to be necessary that everything that
happens in a finite volume of space and time would have to be exactly
analyzable with a finite number of logical operations. The present
theory of physics is not that way, apparently. It allows space to go
down into infinitesimal distances, wavelengths to get infinitely
great, terms to be summed in infinite order, and so forth; and
therefore, if this proposition is right, physical law is wrong.
(Feynman 1981)

\paragraph*{Revise Computer Science.}

Another thing that had been suggested early was that natural laws are
reversible, but that computer rules are not. But this turned out to be
false; the computer rules can be reversible, and it has been a very,
very useful thing to notice and to discover that. This is a place
where the relationship of physics and computation has turned itself
the other way and told us something about the possibilities of
computation. So this is an interesting subject because it tells us
something about computer rules…  (Feynman 1981)

Turing hoped that his abstracted-paper-tape model was so simple, so
transparent and well defined, that it would not depend on any
assumptions about physics that could conceivably be falsified, and
therefore that it could become the basis of an abstract theory of
computation that was independent of the underlying physics. ‘He
thought,’ as Feynman once put it, ‘that he understood paper.’ But he
was mistaken. Real, quantum-mechanical paper is wildly different from
the abstract stuff that the Turing machine uses. The Turing machine is
entirely classical, and does not allow for the possibility the paper
might have different symbols written on it in different universes, and
that those might interfere with one another.  (Deutsch 1985)

Ed Fredkin pursued the idea that information must be finite in
density. One day, he announced that things must be even more simple
than that. He said that he was going to assume that information itself
is conserved. ``You’re out of you mind, Ed.'' I pronounced. ``That’s
completely ridiculous. Nothing could happen in such a world. There
couldn’t even be logical gates. No decisions could ever be made.'' But
when Fredkin gets one of his ideas, he’s quite immune to objections
like that; indeed, they fuel him with energy. Soon he went on to
assume that information processing must also be reversible — and
invented what’s now called the Fredkin gate. (Minsky 1999)

Even revise the laws of thought themselves: In other terms, what is so
good in logic that quantum physics should obey? Can't we imagine that
our conceptions about logic are wrong, so wrong that they are unable
to cope with the quantum miracle? […] Instead of teaching logic to
nature, it is more reasonable to learn from her. Instead of
interpreting quantum into logic, we shall interpret logic into
quantum. (Girard 2007)

@@

A physical principle of computation

No creation of information
No duplication of information

All programs, proofs, deductions are equivalences, isomorphisms,
reversible

Several pages on reversible logic; reversible family of level 1 and
level 2 languages (Pi);

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}

Perhaps Information Flow Security
